---
title: "Stat 3202 Lab 11"
author: "Shivani patel.3999"
date: "3/28/2022"
output:
  pdf_document: default
  html_document: default
---

tinytex::install_tinytex()
# Problem 1

Creating confidence intervals (1-$\alpha$)% for the true population mean difference $\mu_x$-$\mu_y$ and calculating coverage rate

Coverage long term should be equal to 1-$\alpha$

#a)

(1-$\alpha$)% confidence interval for mean difference $\mu_{x}-\mu_{y}$ for two populations X and Y with normal data and known variances $\sigma_{x}^2$ and $\sigma_{y}^2$

$\bar{x}$-$\bar{y}$+-$z_{1-\frac{\alpha}{2}}$* $\sqrt{\frac{\sigma_{x}^2}{x} + \frac{\sigma_{y}^2}{n_{y}}}$ 

#b)

When $\mu_{x}=10$,  $\sigma_{x}=3$, $n_{x}=5$
$\mu_{y}=5$, $\sigma_{y}=5.5$, $n_{y}=8$        $\alpha=.05$ 

Both populations are normal and confidence intervals for the mean difference created and coverage calculated

```{r}
n_x <- 5
n_y <- 8
mu_x <- 10
mu_y <- 5
sigma_x <- 3            
sigma_y <- 5.5
alpha <- .05

CI_dif <- function(n_x, n_y,mu_x, mu_y, sigma_x, sigma_y,alpha){
  sample_x <- rnorm(n_x, mu_x, sigma_x)
  sample_y <- rnorm(n_y, mu_y, sigma_y)
  xbar <- mean(sample_x)
  ybar <- mean(sample_y)
  LL <- xbar-ybar-qnorm(1-alpha/2)*sqrt(sigma_x^2/(n_x) + sigma_y^2/(n_y))
  UL <- xbar-ybar+qnorm(1-alpha/2)*sqrt(sigma_x^2/(n_x) + sigma_y^2/(n_y))
  coverage <- LL<=mu_x-mu_y & UL>=mu_x-mu_y
}

NMC <- 10000
CI <- replicate(NMC, CI_dif(n_x=n_x, n_y=n_y, mu_x=mu_x, mu_y=mu_y ,sigma_x=sigma_x, sigma_y=sigma_y,alpha=alpha))
mean(CI)


```
The coverage rate for the intervals is .954 which is approxomately what 1-$\alpha$ would equal to when $\alpha$=.05


#c)

Confidence intervals calculated when the variances are not known but are equal and we must use the statistic s pooled for $\sigma_{x} = \sigma_{y} = \sigma$

Let $\sigma_{x}=3$ and $\sigma_{y}=3$

i)

$\bar{x}$-$\bar{y}$+-$t_{(1-\frac{\alpha}{2}, n_x+n_y-2)}* s_{p}\sqrt{\frac{1}{n_{x}} + \frac{1}{n_{y}}}$

where $s_{p} = \sqrt{\frac{(n_{x}-1)s_{x}^2+(n_{y}-1)s_{y}^2}{n_{x}+n_{y}-2}}$

```{r}
n_x <- 5
n_y <- 8
mu_x <- 10
mu_y <- 5
sigma_x <- 3            
sigma_y <- 3
alpha <- .05

CI_diff<-function(n_x, n_y, mu_x, mu_y, sigma_x, sigma_y,alpha){
  sample1<-rnorm(n_x,mu_x,sigma_x)
  sample2 <-rnorm(n_y,mu_y,sigma_y)
  xbar<-mean(sample1)
  ybar<-mean(sample2)
  sd_x<-sd(sample1)
  sd_y<-sd(sample2)
  sd_pool2<- (((n_x-1)*sd_x^2 + (n_y-1)*sd_y^2)) / (n_x+n_y-2)
  
  LL<-(xbar-ybar)-qt(1-alpha/2, n_x+n_y-2)*sqrt(sd_pool2)*sqrt(1/n_x + 1/n_y)
  UL<-(xbar-ybar)+qt(1-alpha/2, n_x+n_y-2)*sqrt(sd_pool2)*sqrt(1/n_x + 1/n_y)
  coverage<-mu_x-mu_y>=LL & mu_x-mu_y<=UL
  return(coverage)
}
NMC<-10000
CI<-replicate(NMC,CI_diff(n_x=n_x, n_y=n_y, mu_x=mu_x, mu_y=mu_y ,sigma_x=sigma_x, sigma_y=sigma_y,alpha=alpha))
mean(CI)

```
The coverage rate is once again approximately 1-$\alpha$

ii)
Use t.test function

```{r}
NMC<-10000
coverage<-c()
for(i in 1:NMC){
  sample1<-rnorm(n_x,mu_x,sigma_x)
  sample2<-rnorm(n_y,mu_y,sigma_y)
  out<-t.test(sample1,sample2,var.equal=TRUE, conf.level=1-alpha)
  coverage[i]<-out$conf.int[1]<mu_x-mu_y & out$conf.int[2]>mu_x-mu_y
}
mean(coverage)
```
this value for the coverage rate is even closer to 1-$\alpha$

#Problem 2

Confidence intervals for differences in mean when variance is unknown and unequal
(must use degrees of freedom)

#a)

If v is the degrees of freedom

i)

v = $\frac{(\frac{s_{A}^2}{n_{A}}+\frac{s_{B}^2}{n_{B}})^2}{\frac{s_{A}^4}{n_{A}^2(n_{A}-1)}+\frac{s_{B}^4}{n_{B}^2(n_{B}-1)}}$

and the confidence interval for differences in means is

$\bar{x}$-$\bar{y}$+-$t_{1-\frac{\alpha}{2},v} * \sqrt{\frac{s_{x}^2}{n_{x}}+\frac{s_{y}^2}{n_{y}}}$

ii)

When $\mu_{x}=10$,  $\sigma_{x}=3$, $n_{x}=5$
$\mu_{y}=5$, $\sigma_{y}=5.5$, $n_{y}=8$        $\alpha=.05$ 

```{r}
mu_x <- 10
sigma_x <- 3
n_x <- 5
mu_y <- 5
sigma_y <- 5.5
n_y <- 8
alpha <- .05

NMC<-10000
coverage<-c()
for(i in 1:NMC){
  sample1<-rnorm(n_x,mu_x,sigma_x)
  sample2<-rnorm(n_y,mu_y,sigma_y)
  out<-t.test(sample1,sample2,var.equal=FALSE, conf.level=1-alpha)
  coverage[i]<-out$conf.int[1]<mu_x-mu_y & out$conf.int[2]>mu_x-mu_y
}
mean(coverage)
```
The coverage rate is approximately 1-$\alpha$ when $\alpha=.05$

#Problem 3

Confidence Interval for Difference of Means with Unknown and Unequal Variances but Conservative Degrees of Freedom

degrees of freedom = min($n_{x}-1$, $n_{y}-1$)

so v = min($n_{x}-1$, $n_{y}-1$)

```{r}
mu_x <- 10
sigma_x <- 3
n_x <- 5
mu_y <- 5
sigma_y <- 5.5
n_y <- 8
alpha <- .05
v<-min(n_x-1, n_y-1)

CI_dif<-function(mu_x, sigma_x, n_x, mu_y, sigma_y, n_y, alpha, v){
  sample1<-rnorm(n_x,mu_x,sigma_x)
  sample2<-rnorm(n_y,mu_y,sigma_y)
  xbar<-mean(sample1)
  ybar<-mean(sample2)
  s_x<-sd(sample1)
  s_y<-sd(sample2)
 
  LL<-xbar-ybar-qt(1-alpha/2, v)*sqrt(s_x^2/n_x + s_y^2/n_y)
  UL<-xbar-ybar+qt(1-alpha/2, v)*sqrt(s_x^2/n_x + s_y^2/n_y)
  coverage<-mu_x-mu_y>=LL & mu_x-mu_y<=UL
  return(coverage)
}
NMC<-10000
CI<-replicate(NMC,CI_dif(n_x=n_x, n_y=n_y, mu_x=mu_x, mu_y=mu_y ,sigma_x=sigma_x, sigma_y=sigma_y,alpha=alpha, v=v))
mean(CI)
```
This value is larger than the coverage rate of 1-$\alpha$ because the method of obtaining the degrees of freedom gives the largest margin of error, this is what makes the degrees of freedom conservative.
